An algorithm for a spatial mapping using a hexagon tilegram, with application to Australia disease maps

### 2020-01-10 kobakian_sugarbag_jss/kobakian_sugarbag_jss.tex
JSS version of algorithm paper

---
documentclass: jss
author:
  - name: Stephanie Kobakian
    affiliation: Queensland University of Technology
    address: >
      School of Mthematical Sciences,
      Science and Engineering Faculty,
      Brisbane, QLD, Australia
    email: \email{stephanie.kobakian@qut.edu.au}
    url: http://rstudio.com
title:
  plain: An Algorithm For Spatial Mapping Using a Hexagon Tilegram, With Application to Australian Maps   
abstract: >
  This algorithm creates a tesselated hexagon display to represent each of the spatial polygons. It allocates these hexagon in a manner that preserves the spatial relationship of the geographic units. It showcases spatial distributions, by emphasising the small geographical regions that are often difficult to locate on geographic maps. Spatial distributions have been presented on alternative representations of geography for many years. In modern times, interactivity and animation have begun to play a larger role, as alternative representations have been popularised by online news sites, and atlas websites with a focus on public consumption. Applications are increasingly widespread, especially in the areas of disease mapping, and election results.
keywords:
  # at least one keyword must be supplied
  formatted: [spatial, statistics, cartogram]
preamble: >
  \usepackage{amsmath}
biblio-style: jss
output:
  rticles::jss_article:
    fig_caption: true
    citation_package: natbib
bibliography: sugarbag.bib
link-citations: true
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE, comment = FALSE}
library(knitcitations)
library(RefManageR)
library(sf)

knitr::opts_chunk$set(warning = FALSE, message = FALSE)
options("citation_format" = "pandoc")
BibOptions(check.entries = FALSE, style = "markdown", bib.style = "alphabetic", cite.style = 'alphabetic')
```

# Introduction

<!-- Motivation to map -->

@@

<!-- urbanisation -->
Australian residents are increasingly congregating around major cities, the vast rural areas are often sparsely populated in comparison to the urban centres. 
In Australia, government bodies such as the Australian Bureau of Statistics, and the Australian Electoral Commission hold the responsibility for the division of geographic units. These boundaries are adjusted as the population increases. The division of the population into approximately equal population areas results in dramatically different square meterage of the geographic areas. This results in an unequal attention given to the statistic of each area, this can allow misrepresentation of the spatial distributions of human related statistics in geographic maps.

<!-- solutions -->
The solution to this visualisation problem begins with the geography. Cartograms apply a transformation to the geographic boundaries based on the value of the statistic of interest. These displays result in a distortion of the map space to represent differences in the statistic across the areas \citep{ACCAC}. 
The statistic of interest is used to determine the layout. When using the
Australian population, the result is a population cartogram that fails to preserve a recognisable display due to the difference in size of metropolitan and rural areas \citep{ACTUC}, \citep{GOINO}.
Contiguous cartograms change the shape of an area, preserving boundary relationships of neighbours. Non-contiguous cartograms maintain the geographic shape of each area, but lose the connection to neighbours as each areas shrinks or grows.

@@

# Algorithm

@@

 1. Create the set of centroids to allocate
 2. Create the grid of hexagons locations to use
 3. Allocate each centroid to an available hexagon
 4. Transform the data for plotting

## Parameters

@@

The following must be provided to \code{create_hexmap}:
```{r table_necessary, echo = FALSE, fig.cap = "This table details the parameters that must be provided to create a hexagon tile map using sugarbag."}
library(sugarbag)
library(tidyverse)

knitr::kable(tibble::tibble(parameter = c('shp', 
  'sf_id',
  'focal_points'),
  description = c(
    'an sf object containing the polygon information', 
    'name of a column that distinguishes unique areas',
    'a data frame of reference locations used to allocate hexagons'
    )), format = "latex")
```

## Tasmania

@@

```{r capital_cities}
data(capital_cities)
```

@@

```{r parameters, echo = FALSE, fig.cap = "This table details the additional parameters that are used to create a hexagon tile map using sugarbag."}
knitr::kable(tibble(parameter = c('buffer_dist', 'hex_size',  'hex_filter', 'width'),
  description = c('a float value for distance in degrees to extend beyond the geometry provided',
    'a float value in degrees for the diameter of the hexagons', 'amount of hexagons around centroid to consider for allocation',
    'the angle used to filter the grid points around a centroid')), format = "latex")
```

## Create the set of centroid points

@@

The centroids can be derived from the set of polygons using the \code{create_centroids} function:
```{r cents}
centroids <- create_centroids(shp_sf = tas_sa2, sf_id = "SA2_NAME16")
```

```{r centroids_plot, out.width = "100%", echo = FALSE, eval = FALSE, fig.cap="The geographic shapes of the Statistical Areas of Tasmania at Level 2. The points show the locations of the centroids of the SA2 areas."}
knitr::include_graphics(path = "figures/1centroids.png")
```

## Create the hexagon grid points

@@

```{r grid}
grid <- create_grid(centroids = centroids, hex_size = 0.2, buffer_dist = 1.2)
```

### Step 1: Creating a tesselated grid

@@

```{r grid2, fig.cap="Grid points to create a tilegram.", out.width = "100%", echo = FALSE}
knitr::include_graphics(path = "figures/2grid.png")
```

### Step 2: Rolling windows

@@

### Step 3: Filtering the grid

@@

```{r filter_grid, fig.cap = "All possible hexagon locations from the initial grid are shown with blue outlines. The blue dots shown the grid points left as to choose from after the buffer step. The rolling windows to the right show the rows used to filter the hexagon locations.", echo = FALSE, out.width = "100%"}
knitr::include_graphics(path = "figures/3grid.png", dpi = 300)
```

## Centroid to focal point distance

@@

## Allocate each centroid to a hexagon grid point

@@

```{r centroids, eval = FALSE}
hexmap_allocation <- allocate(
  centroids = centroids %>% select(SA2_NAME16, longitude, latitude),
  sf_id = "SA2_NAME16",
  hex_grid = grid,
  hex_size = 0.2, # same size used in create_grid
  hex_filter = 10,
  width = 35,
  focal_points = capital_cities,
  verbose = TRUE)
```

@@

### Step 2: Filter the grid points for those closest to the centroid

@@

```{r buffers, echo = FALSE, out.width = "100%"}
knitr::include_graphics(path = "figures/4grid.png")
```

@@

If no available hexagon grid point is found within the original filter distance and angle, the distance is expanded, only when a maximum distance is reached will the angle expand to accommodate more possible grid points.  
The angle begins at 30 degrees by default, and will increase if no points can be found within the \code{hex_filter} distance.
The allocation is returned and combined with the data relating to each polygon.

```{r buffs, out.width = "100%", echo = FALSE}
knitr::include_graphics(path = "figures/6allocate.png")
```

@@

# Using sugarbag

```{r full_hexmapcode, echo = TRUE, warning = FALSE, message = FALSE}

# install.packages("sugarbag")
library(sugarbag)

# Create centroids set
centroids <- create_centroids(tas_sa2, "SA2_NAME16")

# Create hexagon location grid
grid <- create_grid(centroids = centroids,
    hex_size = 0.2,
    buffer_dist = 1.2)

# Allocate polygon centroids to hexagon grid points
hex_allocated <- allocate(
  centroids = centroids,
  hex_grid = grid,
  sf_id = "SA2_NAME16",
  # same column used in create_centroids
  hex_size = 0.2,
  # same size used in create_grid
  hex_filter = 10,
  use_neighbours = tas_sa2,
  focal_points = capital_cities %>% filter(points == "Hobart"),
  width = 35,
  verbose = FALSE)
```

## Neighbour relationships

@@

# Australian Cancer Atlas

@@

```{r sir, out.width = "100%", echo = FALSE, fig.cap = "The Australian Cancer Atlas data has determined the colour of each Statistical Area of Australian at Level 2. A choropleth map (a) of Standardised Incidence Rates (SIRs) is paired with a hexagon tile map (b) to contrast the colours that are made obvious when every SA2 is equally represented."}
knitr::include_graphics(path = "figures/7SIR.png")
```

# Conclusion

@@

Alternative maps highlight the value of the statistics across the geographic units. Alternative mapping methods allow increased understanding of the spatial distribution of a variable across the population, by fairly representing each administrative area. This acknowledges that the amount of residents can be different but recognises that each population area is equally important. The solution to this visualisation problem has equally sized areas, with neighbourhood boundary connections. This map algorithm is implemented in the @sugarbag package written for @R.
The \pkg{sugarbag} package creates tesselated hexagon tilegrams. The Australian application preserves the spatial relationships, emphasising capital cities. The hexagon tile map is a visualisation solution that highlights spatial distributions.

>>>>

\documentclass[
]{jss}

\usepackage[utf8]{inputenc}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\author{
Stephanie Kobakian\\Queensland University of Technology
}
\title{}

\Plainauthor{Stephanie Kobakian}
\Plaintitle{An Algorithm For Spatial Mapping Using a Hexagon Tilegram, With
Application to Australian Maps}

\Abstract{
This algorithm creates a tesselated hexagon display to represent each of
the spatial polygons. It allocates these hexagon in a manner that
preserves the spatial relationship of the geographic units. It showcases
spatial distributions, by emphasising the small geographical regions
that are often difficult to locate on geographic maps. Spatial
distributions have been presented on alternative representations of
geography for many years. In modern times, interactivity and animation
have begun to play a larger role, as alternative representations have
been popularised by online news sites, and atlas websites with a focus
on public consumption. Applications are increasingly widespread,
especially in the areas of disease mapping, and election results.
}

\Keywords{spatial, statistics, cartogram}

%% publication information
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{}
%% \Acceptdate{2012-06-04}

\Address{
    Stephanie Kobakian\\
  Queensland University of Technology\\
  School of Mthematical Sciences, Science and Engineering Faculty,
  Brisbane, QLD, Australia\\
  E-mail: \email{stephanie.kobakian@qut.edu.au}\\
  URL: \url{http://rstudio.com}\\~\\
  }

% Pandoc header

\usepackage{amsmath}

\begin{document}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

Australian residents are increasingly congregating around major cities,
the vast rural areas are often sparsely populated in comparison to the
urban centres. In Australia, government bodies such as the Australian
Bureau of Statistics, and the Australian Electoral Commission hold the
responsibility for the division of geographic units. These boundaries
are adjusted as the population increases. The division of the population
into approximately equal population areas results in dramatically
different square meterage of the geographic areas. This results in an
unequal attention given to the statistic of each area, this can allow
misrepresentation of the spatial distributions of human related
statistics in geographic maps.

The solution to this visualisation problem begins with the geography.
Cartograms apply a transformation to the geographic boundaries based on
the value of the statistic of interest. These displays result in a
distortion of the map space to represent differences in the statistic
across the areas \citep{ACCAC}. The statistic of interest is used to
determine the layout. When using the Australian population, the result
is a population cartogram that fails to preserve a recognisable display
due to the difference in size of metropolitan and rural areas
\citep{ACTUC}, \citep{GOINO}. Contiguous cartograms change the shape of
an area, preserving boundary relationships of neighbours. Non-contiguous
cartograms maintain the geographic shape of each area, but lose the
connection to neighbours as each areas shrinks or grows.

@@

\hypertarget{algorithm}{%
\section{Algorithm}\label{algorithm}}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Create the set of centroids to allocate
\item
  Create the grid of hexagons locations to use
\item
  Allocate each centroid to an available hexagon
\item
  Transform the data for plotting
\end{enumerate}

\hypertarget{parameters}{%
\subsection{Parameters}\label{parameters}}

@@

The following must be provided to \code{create_hexmap}:

\begin{CodeChunk}

\begin{tabular}{l|l}
\hline
parameter & description\\
\hline
shp & an sf object containing the polygon information\\
\hline
sf\_id & name of a column that distinguishes unique areas\\
\hline
focal\_points & a data frame of reference locations used to allocate hexagons\\
\hline
\end{tabular}

\end{CodeChunk}

\hypertarget{tasmania}{%
\subsection{Tasmania}\label{tasmania}}

@@

\begin{CodeChunk}

\begin{CodeInput}
R> data(capital_cities)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}

\begin{tabular}{l|l}
\hline
parameter & description\\
\hline
buffer\_dist & a float value for distance in degrees to extend beyond the geometry provided\\
\hline
hex\_size & a float value in degrees for the diameter of the hexagons\\
\hline
hex\_filter & amount of hexagons around centroid to consider for allocation\\
\hline
width & the angle used to filter the grid points around a centroid\\
\hline
\end{tabular}

\end{CodeChunk}

\hypertarget{create-the-set-of-centroid-points}{%
\subsection{Create the set of centroid
points}\label{create-the-set-of-centroid-points}}

@@

The centroids can be derived from the set of polygons using the
\code{create_centroids} function:

\begin{CodeChunk}

\begin{CodeInput}
R> centroids <- create_centroids(shp_sf = tas_sa2, sf_id = "SA2_NAME16")
\end{CodeInput}
\end{CodeChunk}

\hypertarget{create-the-hexagon-grid-points}{%
\subsection{Create the hexagon grid
points}\label{create-the-hexagon-grid-points}}

@@

\begin{CodeChunk}

\begin{CodeInput}
R> grid <- create_grid(centroids = centroids, hex_size = 0.2, buffer_dist = 1.2)
\end{CodeInput}
\end{CodeChunk}

\hypertarget{step-1-creating-a-tesselated-grid}{%
\subsubsection{Step 1: Creating a tesselated
grid}\label{step-1-creating-a-tesselated-grid}}

@@

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{figures/2grid}

}

\caption[Grid points to create a tilegram]{Grid points to create a tilegram.}\label{fig:grid2}
\end{figure}
\end{CodeChunk}

\hypertarget{step-2-rolling-windows}{%
\subsubsection{Step 2: Rolling windows}\label{step-2-rolling-windows}}

@@

\hypertarget{step-3-filtering-the-grid}{%
\subsubsection{Step 3: Filtering the
grid}\label{step-3-filtering-the-grid}}

@@

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{figures/3grid}

}

\caption[All possible hexagon locations from the initial grid are shown with blue outlines]{All possible hexagon locations from the initial grid are shown with blue outlines. The blue dots shown the grid points left as to choose from after the buffer step. The rolling windows to the right show the rows used to filter the hexagon locations.}\label{fig:filter_grid}
\end{figure}
\end{CodeChunk}

\hypertarget{centroid-to-focal-point-distance}{%
\subsection{Centroid to focal point
distance}\label{centroid-to-focal-point-distance}}

@@

\hypertarget{allocate-each-centroid-to-a-hexagon-grid-point}{%
\subsection{Allocate each centroid to a hexagon grid
point}\label{allocate-each-centroid-to-a-hexagon-grid-point}}

@@

\begin{CodeChunk}

\begin{CodeInput}
R> hexmap_allocation <- allocate(
R>   centroids = centroids %>% select(SA2_NAME16, longitude, latitude),
R>   sf_id = "SA2_NAME16",
R>   hex_grid = grid,
R>   hex_size = 0.2, # same size used in create_grid
R>   hex_filter = 10,
R>   width = 35,
R>   focal_points = capital_cities,
R>   verbose = TRUE)
\end{CodeInput}
\end{CodeChunk}

@@

\hypertarget{step-2-filter-the-grid-points-for-those-closest-to-the-centroid}{%
\subsubsection{Step 2: Filter the grid points for those closest to the
centroid}\label{step-2-filter-the-grid-points-for-those-closest-to-the-centroid}}

@@

\begin{CodeChunk}

\begin{center}\includegraphics[width=1\linewidth]{figures/4grid} \end{center}

\end{CodeChunk}

@@

If no available hexagon grid point is found within the original filter
distance and angle, the distance is expanded, only when a maximum
distance is reached will the angle expand to accommodate more possible
grid points.\\
The angle begins at 30 degrees by default, and will increase if no
points can be found within the \code{hex_filter} distance. The
allocation is returned and combined with the data relating to each
polygon.

\begin{CodeChunk}

\begin{center}\includegraphics[width=1\linewidth]{figures/6allocate} \end{center}

\end{CodeChunk}

@@

\hypertarget{using-sugarbag}{%
\section{Using sugarbag}\label{using-sugarbag}}

\begin{CodeChunk}

\begin{CodeInput}
R> # install.packages("sugarbag")
R> library(sugarbag)
R> 
R> # Create centroids set
R> centroids <- create_centroids(tas_sa2, "SA2_NAME16")
R> 
R> # Create hexagon location grid
R> grid <- create_grid(centroids = centroids,
R+     hex_size = 0.2,
R+     buffer_dist = 1.2)
R> 
R> # Allocate polygon centroids to hexagon grid points
R> hex_allocated <- allocate(
R+   centroids = centroids,
R+   hex_grid = grid,
R+   sf_id = "SA2_NAME16",
R+   # same column used in create_centroids
R+   hex_size = 0.2,
R+   # same size used in create_grid
R+   hex_filter = 10,
R+   use_neighbours = tas_sa2,
R+   focal_points = capital_cities %>% filter(points == "Hobart"),
R+   width = 35,
R+   verbose = FALSE)
\end{CodeInput}
\end{CodeChunk}

\hypertarget{neighbour-relationships}{%
\subsection{Neighbour relationships}\label{neighbour-relationships}}

@@

\hypertarget{australian-cancer-atlas}{%
\section{Australian Cancer Atlas}\label{australian-cancer-atlas}}

@@

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{figures/7SIR}

}

\caption[The Australian Cancer Atlas data has determined the colour of each Statistical Area of Australian at Level 2]{The Australian Cancer Atlas data has determined the colour of each Statistical Area of Australian at Level 2. A choropleth map (a) of Standardised Incidence Rates (SIRs) is paired with a hexagon tile map (b) to contrast the colours that are made obvious when every SA2 is equally represented.}\label{fig:sir}
\end{figure}
\end{CodeChunk}

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

@@

Alternative maps highlight the value of the statistics across the
geographic units. Alternative mapping methods allow increased
understanding of the spatial distribution of a variable across the
population, by fairly representing each administrative area. This
acknowledges that the amount of residents can be different but
recognises that each population area is equally important. The solution
to this visualisation problem has equally sized areas, with
neighbourhood boundary connections. This map algorithm is implemented in
the \citet{sugarbag} package written for \citet{R}. The \pkg{sugarbag}
package creates tesselated hexagon tilegrams. The Australian application
preserves the spatial relationships, emphasising capital cities. The
hexagon tile map is a visualisation solution that highlights spatial
distributions.

\bibliography{sugarbag.bib}

\end{document}
